---
title: "Introduction to `noreden`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to `noreden`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`noreden` provides user-friendly functions to facilitate sustainable diet discovery. Given information for food (groups) and their nutritional and environmental impact information, the aim is to find a new diet (i.e. set of foods) that is healthy, acceptable, yet more sustainable in terms of environmental impact. 

The optimization problem implemented minimizes the **difference between the current and new target diet**, subject to a set of inequality constraints with respect to nutrition and environmental impact. More details can be found in this document: [Optimization for NorEden](https://ocbe-uio.github.io/noreden_doc/documentation/).

In order to find a new diet, you need the following input data: 

* **current diet**: food group names and intake (gram) for each food group
* **nutrition (per unit)**: used to compute total nutrition, such as energy, protein.
* **environment impact (per unit)**: used to compute total impact, such as ghge (green house gas emission)
* **constraint**: we implement two inequality constraints for each nutrition / env impact, such that the total amount ('contribution') falls within a range. In this package the values are based on current diet's total contribution.

# An example

We use a simple example to demonstrate the main features of the package. To get started, load the package.

```{r setup}
library(noreden)
```



## Process input

We need some information on the current diet we wish to optimize. For now, we use the **average diet** summarized from a Norwegian dietary survey. However in the future, we shall implement flexible input options so that you can insert the data yourself.

```{r}
head(all_diet)
```

We also need the **contribution per unit** (referred to as `cpu`) for each of the food. The food unit is per gram, and units for nutrition and environmental impact depends on the outcome. 

```{r}
head(contrib_per_unit)
```


### Select food and outcomes of interest

We select 12 foods, and 5 outcomes of interest.

```{r}
tag_food_12 <- c('Bread', 'Vegetables', 'Red meat', 
                'Milk, yoghurt', 'Fish', 'Cheese', 
                'Eggs', 'Fruit, berries', 'Potatoes', 
                'Other grains', 'Butter, margarine, oil', 'Sugar, sweets')

tag_outcome_5 <- c('energy', 'protein', 'carbs', 'fat', 'ghge')
```

For the current diet selection, we get the average diet (`intake_mean`) as a baseline, from which we search for a new diet that satisfy our requirements; in addition, we also get the upper and lower bound for the new diet so that it is not too far from the current one. 

```{r}
diet_selected <- select_diet(
  data_diet = all_diet,
  tag_food = tag_food_12)

diet_selected
```

Do the same for outcomes on selected foods.

```{r}
cpu_selected <- select_perunit(
  data_perunit_contrib = contrib_per_unit, 
  tag_food = tag_food_12, 
  tag_outcome = tag_outcome_5)
cpu_selected
```


## Compute total contribution 

We compute the total contribution of the food combination. This is the basis when we search for new diet that satisfies the nutrition and environmental constraints.

```{r}
tc <- compute_total_contrib(
  data_diet = diet_selected, 
  data_perunit_contrib = cpu_selected)

tc
```


## Set constraints 

The inequality constraints look like this:

* total energy <= upper bound for energy
* total energy >= lower bound for energy

By default we can set the upper bound to be the same as the current total outputs, and lower bound to be 90% of the current total outputs.


```{r}
constr_coef_df <- set_constr_coef(
  tag_outcome = tag_outcome_5, 
  coef_lwr = rep(0.9, length(tag_outcome_5)), 
  coef_upr = rep(1.0, length(tag_outcome_5)))
constr_coef_df
```


When we try to reduce the output in a specific outcome (such as ghge), we can set it to be 90% the previous level: the function `reduce_constr()` allows you to set it.

```{r}
# reduce ghge to 0.9
constr_coef_df_red <- reduce_constr(
  data_constr_coef = constr_coef_df, 
  tag_outcome_reduce = 'ghge', 
  coef_reduce = 0.9)
constr_coef_df_red
```

Use total contribution and constraint coefficients to compute the upper and lower bound.

```{r}
constr_val_reduce <- compute_constr(
  data_total_contrib = tc$total_contrib, 
  data_constr_coef = constr_coef_df_red)
constr_val_reduce
```


### Set standardized constraints

Sometimes it is better for computaion to use standardized values rather than the original. We have implemented one way to standardize.

```{r}
stdcoef <- compute_stdcoef(data_perunit_contrib = cpu_selected)
stdcoef

cpu_selected_std_res <- compute_std_unit_contrib(
  uc_raw = cpu_selected,
  std_coef = stdcoef$std_coef)

cpu_selected_std <- cpu_selected_std_res$uc_std

# total contrib (std)
tc_std <- compute_total_contrib(
  data_diet = diet_selected, 
  data_perunit_contrib = cpu_selected_std)
tc_std
```


Now we directly use the 90% ghge constraint coefficients.

```{r}
# constraints based on std total contrib
constr_val_reduce_std <- compute_constr(
  data_total_contrib = tc_std$total_contrib, 
  data_constr_coef = constr_coef_df_red)

constr_val_reduce_std
```



## Find new diet

In order to run the algorithm to search for new diet, we need to provide the algorithm the data we prepared from previous steps in a format that is ordered by outcome.

Here we use the standardized constraints for numerical stability.

```{r}
# split constraint values 
constval <- values_by_tag_outcome(
  data_unit_contrib = cpu_selected_std, 
  data_constr = constr_val_reduce_std)

constval$food_name
constval$tag_outcome

# for energy, unit_contrib per food, lwr and upr bound of total contrib
constval$val$energy
```


Run the optimization, and retrieve the new diet along with current diet.

```{r}
res <- find_new_diet(diet0 = diet_selected$intake_mean, 
                     diet0_upr = diet_selected$intake_upr, 
                     diet0_lwr = diet_selected$intake_lwr, 
                     tag_outcomes = tag_outcome_5, 
                     constraint_val = constval$val, 
                     print_runtime = T)
# collect result
new_diet <- return_new_diet(
  result_obj = res$run_optim, 
  data_current_diet = diet_selected)
new_diet
```


## Present the results

We can show some useful comparisons between the two diets, in terms of absolute and relative percent change.

```{r}
# compute difference
new_old_compare <- compare_new_diet(data_new_diet = new_diet, 
                 data_current_diet = diet_selected)

new_old_compare
```


We can also validate whether the new diet satisfies the constraints.

```{r}
# validate constraints
new_diet_validate <- validate_diet_contrib(data_new_diet = new_diet, 
                      data_unit_contrib = cpu_selected,
                      data_constr = constr_val_reduce)

new_diet_validate
```


It is possible to visualize the results or present results in better looking tables. 

```{r}
d_compare_gram <- prep_diet_comparison_gram(
  new_old_compare)

p1 <- plot_diet_comparison_gram(plot_obj = d_compare_gram,
                          title_text = 'New diet',
                          axis_x_text = 'Food groups',
                          axis_y_text = 'Intake (grams)')
p1
```

Table with comments.

```{r}
d_tb_validate <- prep_validate_table(
  data_validate_diet = new_diet_validate)
table_validate(d_tb_validate)
```

